---
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
---

# Writing functions in R

## What is a function?

Functions are a way to package up bits of code to make them easy to reuse. Base R includes numerous built-in functions and there are thousands more R functions in packages available on CRAN and elsewhere.


Here is an example of one of the built-in R functions, `sum()`:

```{r}

x <- c(1, 2, 3, 4, 5) # Create a vector of numbers to sum
sum_x <- sum(x) # Sum the numbers contained in 'x', and save the result as a new variable 'sum_x'

```

You can also write your own functions, called "user-defined functions".

Functions in a package are the same thing as functions you define yourself, they're just stored in a different way.

You can see the code underlying a function by typing its name (without brackets) in the console and hitting "enter".

## Why use functions?

### Code maintenance

Generally functions are used to automate common tasks, to avoid copying and pasting the same code multiple times with minor alterations. While writing a function has a small up front cost, you'll reap the benefits when you need to make changes to your code, and your QA'er will also thank you, as they'll only need to check your function works once!

### Abstraction

One of the benefits of functions is they abstract away the details of *how* the code works. To use a function, all you need to understand is *what* the function is designed to do. You'll need to understand how your function works when you're writing it of course, but you won't need to think about it everytime you use it.

### Code legibility

You can use functions to make your code better structured and easier to read - done well, this can make your code a lot easier to understand for someone unfamiliar with it, or even yourself a few months down the line.


# How to write a function

## Structure of a function

The syntax for creating a function is:

```{r}
function_name <- function(arg1, arg2){
        # function body, e.g. 
        print(paste(arg1, arg2))
}

```

We use `function()` to create a function and assign it to the object `function_name`.

A function is made up of three components:

1. Its **arguments** (in this example, `arg1` and `arg2`) - these are variables used inside the function body which we can set each time we call the function.
   
2. The function **body** (everything between the curly brackets) - this is where we put the code.
   
3. And the function **environment** (where the function was created) - this determines what variables and other functions it has access to - you can find out more about environments  [here](https://r4ds.had.co.nz/functions.html#environment).

# Examples of functions

## Example 1: a basic function

Here's an example of a very basic user-defined function, that adds two to a numerical variable:

```{r}

# we've named our function 'add_two' and have included 'x' as an input variable (a function argument)
add_two <- function(x){
  # This function takes the argument 'x', and adds 2 to it
  result <- x + 2
  return(result) # The output of the function
}

```

Let's try using the function `add_two()`:

```{r}
# Calling the function by itself prints the result to the console
add_two(3)

# Alternatively the result can be saved as a new variable
result <- add_two(3) 

```

---

The return statement is not always necessary, as R will automatically return the variable on the last line of the function body. So this version of the function also works:

```{r}

add_two <- function(x){
  result <- x + 2
}

result <- add_two(3)

```

## Example 2: function inputs & outputs

Functions can accept as many arguments as you like, but can only output one object (if it is necessary to output more than one object, then the desired outputs can be combined together into a list, and the list can be returned).

```{r}

sum_squares <- function(x, y){
  result <- x^2 + y^2
  return(result)
}

z <- sum_squares(3, 5)

```

Arguments can be specified by position or name. In the above example, the function accepts arguments called `x` and `y`, with the order specified as `x` then `y` in the function definition. Therefore when `sum_squares(3, 5)` is called, the arguments are interpreted as `x = 3` and `y = 5`. Alternatively, if we called `sum_squares(5, 3)` then the arguments would be interpreted as `x = 5` and `y = 3`.

For more complicated functions with lots of arguments, specifying the arguments by position becomes more error-prone. Therefore it is often desirable to specify the arguments by name, then the order of the arguments doesn't matter:

```{r}

sum_squares(x = 3, y = 5)
sum_squares(y = 5, x = 3)

```

## Example 3: using return()

Functions can return different outputs depending on some condition, using the `return()` command:

```{r}

mod_x <- function(x){
  
  if(x < 0){
    return(-x)
  }
  else{
    return(x)
  }
  
}

x <- -5
mod_x(x)

```

As soon as a `return()` statement is encountered in a function, the function finishes and returns the variable in the `return()` statement. In the above example, `return()` is used twice as there is a condition with two possibilities. Either the input argument `x` is less than zero, in which case the function ends with the `return(-x)` statement, otherwise if `x` is not less than zero then the function continues to the `return(x)` statement.

## Example 4

Depending on the purpose of the function, it doesn't need to return an output at all:

```{r}

odd_or_even <- function(x){
  
  if((x %% 2) == 0){
    print("The number is even.")
  }
  else if((x %% 2) == 1) {
    print("The number is odd.")
  }

}

odd_or_even(x=4)

```

## Example 5: errors and warnings

Sometimes it can be useful to include helpful error messages in functions, by anticipating the sorts of variables that could input. What happens if you try to use a non-integer or a string as an argument in the above example?

```{r error = TRUE}
odd_or_even(x=1.5)
odd_or_even(x="a")
odd_or_even(x=c(1,2,3))

```

---

Here is an adapted version of the function, with some more informative error messages built in:

```{r}

odd_or_even <- function(x){
  
  if (length(x) > 1){
    stop("The argument is not a single number.")
  }
  else if (!is.numeric(x)){
    stop("The argument is not a number.")
  }
  else if((x %% 2) == 0){
    print("The number is even.")
  }
  else if((x %% 2) == 1) {
    print("The number is odd.")
  }
  else{
    stop("The number is not an integer.")
  }

}

```

---

Now try passing some incompatible arguments to the function:

```{r error = TRUE}
odd_or_even(x=1.5)
odd_or_even(x="a")
odd_or_even(x=c(1,2,3))

```

`stop()` halts execution of the function and prints an error message to the console; alternatively you can use `warning()` which prints a warning message to the console but does not stop execution of the function.

## Example 6: optional arguments

Here's an example of how to include optional arguments:

```{r}

add_a_number <- function(x, y = NULL){ # y is the optional input, with a default value of NULL
  
  if(!is.null(y)){
    return(x + y)
  }
  else{
    return(x)
  }
  
}

add_a_number(x = 6)
add_a_number(x = 6, y = 7)

```

## Example 7: arguments with default values

Any value can be used as a default value for an argument. For example, we can generalise the sum_squares function above by allowing it to sum together two numbers raised to any power, but have a default power of 2:

```{r}

sum_powers <- function(x, y, z = 2){
  
  result <- x ^ z + y ^ z
  return(result)
  
}

sum_powers(x = 3, y = 5)
sum_powers(x = 3, y = 5, z = 3)

```

## Example 8: the ellipsis argument

Sometimes being able to pass an arbitrary number of arguments can be useful, especially when another function is called within a wrapper function. This requires the ellipsis construct `...`, which is designed to pass a variable number of arguments to a function. Here's an example:

```{r}

plot_x_and_y <- function(x, y, ...){
  
  plot(x, y, ...)
  
}

x <- 1:10
y <- (1:10) * 2

```

---

The function can be called with only the required arguments:

```{r}

plot_x_and_y(x, y)

```

---

Or the function can be called with any optional arguments accepted by the plot function:

```{r}

plot_x_and_y(x, y, col='red', type='l')

```

## Example 9

Here's a less trivial example - this function takes a vector of strings as the argument, ensures all letters in the string array are lower-case, and makes use of "regular expressions" or "regex" to find and replace all spaces and punctuation marks with an underscore:

```{r}

generalise_names <- function(names){
  #library(stringr)
  names <- tolower(names) # Convert any uppercase letters to lowercase
  names <- stringr::str_trim(names) # Trim any blank spaces at the start or end of each string
  names <- stringr::str_replace_all(names, " ", "_") # Replace spaces with an underscore
  names <- stringr::str_replace_all(names, "[^A-Za-z0-9]", "_") # Replace anything that isn't a letter or number with an underscore
  names <- stringr::str_replace_all(names, "([_])\\1+", "\\1") # Replace repeated underscores with a single underscore
  names <- stringr::str_replace_all(names, "^\\_+|\\_+$", "") # Remove any underscores at the beginning or end of the string
  
  return(names)
}

names <- c("Option 1", "   Option (1)", "Option: 1", "option 1", "OPTION - 1")

generalise_names(names)

```

## When to write a function

### When you've copied and pasted three times
There is a principal in software engineering called Don't Repeat Yourself (DRY) - which basically states that you should avoid duplication wherever possible. A good rule of thumb is whenever you find you've copied and pasted the same code three times, it's time to consider replacing that code with a function.

### To structure your code
You may also sometimes want to write a function for code you're only planning to use once as a way of structuring your code and "hiding" some of it to make your main script easier to read.

### When someone hasn't already written one for you
The R ecosystem is full of high quality packages designed to solve all kinds of problems - it's generally best to make sure that a function doesn't already exist before writing your own.

## Best practice
Writing a function is easy, writing a really good function can be a lot harder! Here are a few things to consider:

### Give your function a good name
The name of a function should give you a good idea of what it does. Generally function names should be concise and use verbs rather than nouns.

### Document your code
You should have comments explaining what your function does, what each argument is, and what it returns.

### Keep it short
A rule of thumb is if all the code for your function doesn't fit on your screen at the same time, it's probably too complicated. Consider splitting it up into multiple functions.

---

### Generalise
Think about whether there are ways you can make your function usable in more situations. For example, is there anything you're hard-coding which you could set as an argument instead? Don't get carried away though, make sure you don't add complicated functionality you're never going to use!

### Vectorise by default
R is designed to work well with vectors (e.g. columns of a dataframe). Where possible you should write your function so it can take a vector as input and apply the transformation to each element. The `generalise_names()` function we looked at is a good example of this!

## How to organise your code

Whenever you're working on something in R it's generally best to create an R project and version control your code on GitHub. There's information on how to do this in the Analytical Platform guidance.

It's also best to keep your functions separate from the rest of your code to make them easier to find.

### Storing your functions in your project

The easiest way to store your functions is just to create a folder in your project called "functions" and save your functions there.

You could either put each function in its own R script with the same name, or you could group related functions into clearly named scripts.

Then just use `source("my_functions_script.R")` (with `"my_functions_script.R"` replaced with the name of your script) to run the code and make your functions available to you in the current session. As with loading libraries, it's best to do this at the top of your script.

## Writing a package
An alternative is to make your own package to store your functions which you can then use like any other R package. There are a few advantages to this:

+ It means you (and others) can access your functions from different projects
+ There are certain requirements for making R packages which enforce good practice, such as including documentation

This comes at the cost of slightly higher overheads.

A few examples of packages written by people in MoJ are:

+ [s3tools](https://github.com/moj-analytical-services/s3tools)
+ [asdrap](https://github.com/moj-analytical-services/asdrap)

To find out more about writing packages, check out the links under further reading.

## Further reading

### On functions
+ [DataCamp writing functions in R course](https://www.datacamp.com/courses/writing-functions-in-r)
+ [Functions chapter](https://adv-r.hadley.nz/functions.html) of Hadley's Advanced R book
+ [Tidy Evaluation](https://tidyeval.tidyverse.org/introduction.html) (useful for writing functions which behave like Tidyverse functions) 

### On packages
+ [Writing an R package from scratch](https://hilaryparker.com/2014/04/29/writing-an-r-package-from-scratch/)
+ Hadley's [R packages](http://r-pkgs.had.co.nz/) book

### Misc
+ [Tidyverse style guide](https://style.tidyverse.org/) (has some guidance on choosing function and argument names)
+ [Scoping/environments](https://bookdown.org/rdpeng/rprogdatascience/scoping-rules-of-r.html)


