---
output: rmarkdown::github_document
---

# Appendix

## Table of operators

This table shows some of the logical operators you are likely to encounter when using R.

| Operator | Definition |
| :------: | :--------- |
|    ==    | Equal to                      |
|    !=    | Not equal to                  |
|    >     | Greater than                  |
|    <     | Less than                     |
|    >=    | Greater than or equal to      |
|    <=    | Less than or equal to         |
|  &#448;  | Or                            |
|    &     | And                           |
|    !     | Not                           |
|   %in%   | The subject appears in a list |
| is.na()  | The subject is NA             |

## Additional NSE examples

Below are more examples of how to handle `dplyr()`'s non-standard evaluation. If you wanted to use the `group_by()` and `summarise()` functions in a user-defined function, then one option is to have variables containing the column names as strings, and including these variables as function arguments. In the function these variables can be enclosed by `!!as.name()` to convert them into names, like so:

```{r}
sum_group_alt1 <- function(df, group_cols, sum_col) {

  summary <- df %>%
    dplyr::group_by(!!as.name(group_cols)) %>%
    dplyr::summarise(counts = sum(!!as.name(sum_col)))

  return(summary)

}

prosecutions_grouped <- sum_group_alt1(df = prosecutions, group_cols = "age_range", sum_col = "count")
prosecutions_grouped
```

Alternatively, this version of the function means the column names can be input as function arguments directly (rather than needing to enclose them in quote marks to turn them into strings).

```{r}
sum_group_alt2 <- function(df, group_cols, sum_col) {

  summary <- df %>%
    dplyr::group_by(!!enquo(group_cols)) %>%
    dplyr::summarise(counts = sum(!!enquo(sum_col)))

  return(summary)

}

prosecutions_grouped <- sum_group_alt2(df = prosecutions, group_cols = age_range, sum_col = count)
prosecutions_grouped
```

The function below shows an example of how our `sum_group()` function can be modified to accept column names as arguments, using the `{{` operator. Note that you need version 0.4 or later of the rlang package to use the `{{` operator.

```{r}
sum_group_alt3 <- function(df, group_cols, sum_col) {

  summary <- df %>%
    dplyr::group_by({{ group_cols }}) %>%
    dplyr::summarise(counts = sum({{ sum_col }}))

  return(summary)
  
}

prosecutions_grouped <- sum_group_alt3(df = prosecutions, group_cols = age_range, sum_col = count)
prosecutions_grouped
```

## Bonus examples

### Adding an optional total row

We can extend the `sum_group()` function by having the option to add a row with the total across all categories. Note that this requires the `janitor` package.

```{r, message = FALSE}
library(janitor)

sum_group <- function(df, group_cols, sum_col, add_total=F) {
  
  summary <- df %>%
    dplyr::group_by_at(group_cols) %>%
    dplyr::summarise_at(sum_col, sum)
  
  if (add_total == T) {
    summary <- summary %>% janitor::adorn_totals("row")
  }
  
  return(summary)
  
}
```

### Extracting a subset of the data

Sometimes processing data requires manipulating dates and times. For example, if we wanted to extract the prosecutions from the year up to a particular date, we could use a function like:

```{r}
# This function extracts the prosecutions from a particular year
extract_year <- function(data, end_date) {
  
  # Ensure the date is a date-time object
  if (is.character(end_date)) { end_date <- lubridate::dmy(end_date) }
  
  # Find end of quarter dates for the past year
  quarters_to_include <- end_date %m-% months(c(0, 3, 6, 9))
  
  # Format the dates to years and quarters
  years <- lubridate::year(quarters_to_include)
  quarters <- quarters(quarters_to_include)

  # Combine into a unique set of year-quarters
  yearquarters <- str_c(years, " ", quarters)
  
  # Filter data based on these years and quarters
  data <- data %>%
    dplyr::mutate(year_quarter = paste(year, quarter)) %>%
    dplyr::filter(year_quarter %in% yearquarters)
  
  return(data)
}
```

```{r}
prosecutions_extract <- extract_year(prosecutions, "31-Mar-2018")
glimpse(prosecutions_extract)
```
