---
title: Writing functions in R
output: rmarkdown::github_document
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = "../",
      output_file = file.path("../README.md")  
    )
  })
---

```{r setup, include=FALSE}
# These variables determine whether or not exercise solutions are included
show_solution <- FALSE # This determines if the solutions are displayed in the readme
purl_solutions <- FALSE # This variable relates to code blocks that are exercise solutions
purl_example_code <- TRUE # This variable relates to code blocks that aren't exercise solutions
```

This repository is for the Writing Functions in R course offered by the Data and Analysis R Training Group. 

Knowing how to write your own functions is a great skill to add to your R toolbox. Writing functions can save you time, reduce the risk of errors, and make your code easier to understand. In this course we’ll cover why, when and how to write your own functions, with plenty of examples & exercises to help you get started.

The session is intended to be accessible to anyone who is familiar with the content of the [Introduction to R](https://github.com/moj-analytical-services/IntroRTraining) training session & has some experience of using R in their work.

This training session is periodically run in person/over Teams. Alternatively, you can go through this material in your own time - all the notes are available below and you can also find the recording of a previous session [here](https://justiceuk.sharepoint.com/:f:/s/RTrainingGroup/Ev1kwXxGaVNNpLwFg4drakUBNUpZz72DRdcD8PipkU-u9A?e=SQzeJL). See [Remote learning](#remote-learning) for more tips on going through this material in your own time. If you work through the material by yourself please leave feedback about the material [here](https://airtable.com/shr9u2OJB2pW8Y0Af).

If you have any feedback on the content, please get in touch!

## Contents

* [Pre-material](#pre-material)
* [Remote learning](#remote-learning)
* [Learning outcomes](#learning-outcomes)
* [What is a function?](#what-is-a-function)
* [Why use functions?](#why-use-functions)
* [How to write a function](#how-to-write-a-function)
* [Examples of basic functions](#examples-of-basic-functions)
* ["Real-world" example functions](#real-world-example-functions)
* [When to write a function](#when-to-write-a-function)
* [Best practice](#best-practice)
* [How to organise your code](#how-to-organise-your-code)
* [Writing a package](#writing-a-package)
* [Further reading](#further-reading)
* [Appendix](#appendix)
* [Information for presenters](#information-for-presenters)

## Pre-material

A few days before the session, please make sure that -

1. You have access to RStudio on the Analytical Platform
2. You have requested access to the **alpha-r-training** s3 bucket via the [intro_r slack channel on the ASD
workspace](https://moj.enterprise.slack.com/archives/CGKSJV9HN)
3. You have followed the steps in the [Configure Git and Github section of the Platform User Guidance](https://user-guidance.analytical-platform.service.justice.gov.uk/github/set-up-github.html) to configure Git and GitHub (this only needs doing once)
4. You have cloned this repository (instructions are in the Analytical Platform User Guidance [here](https://user-guidance.analytical-platform.service.justice.gov.uk/github/rstudio-git.html#work-with-git-in-rstudio))
5. You run the command `renv::restore()` in the console to make sure you have the required packages installed

If you have any problems with the above please get in touch with the course organisers or ask for help on either the #analytical_platform or #intro_r channel on [ASD slack](https://asdslack.slack.com).

All the examples in the presentation/README are available in the R script example_code.R.

## Remote learning

Here are a few suggestions if you are going through this material in your own time:

* Both the README and slides contain the same content so you can use whichever works best for you - the only difference is the exercise solutions are only in the slides
*  To open the slides, first clone the repo (see [pre-material](#pre-material)) and then open the file "slides.html" from RStudio by clicking "View in web browser"
* You can find a recording of a previous training session on the R training Microsoft Stream channel [here](https://web.microsoftstream.com/channel/aa3cda5d-99d6-4e9d-ac5e-6548dd55f52a)
* If you need any help, the best place to ask is on either the [#intro_r](https://app.slack.com/client/T1PU1AP6D/CGKSJV9HN) or [#r](https://app.slack.com/client/T1PU1AP6D/C1PUCG719) slack channels on ASD slack.

If you have any feedback on the material and/or your experience of working through it in your own time, please get in touch with a member of the R training group or leave a review on [Airtable](https://airtable.com/shr9u2OJB2pW8Y0Af).


```{r source_content, child = 'content.Rmd'}
```

```{r source_content, child = 'appendix.Rmd'}
```

## Information for presenters

### Where stuff is for on the day
* The training slides are in [slides.html](slides.html) (you will need to download these and open them in a browser).
* The course material is also duplicated in the README with some additional examples in the appendix.
* The code from the presentation is available in [example_code.R](example_code.R).
* The answers to the exercises are available in [solutions.R](solutions.R).
* The [functions.R](functions.R) script is used in one example to demonstrate how to organize code.

### How to update the course contents
The README, slides, solutions, and example_code are all generated from files in the "rmd_files" folder.

* The bulk of the content is in [content.Rmd](rmd_files/content.Rmd).
* The appendix (used in the README) is in [appendix.Rmd](rmd_files/appendix.Rmd).
* The README is generated from [README.Rmd](rmd_files/README.Rmd) (this will source the contents of
content.Rmd and appendix.Rmd).
* The slides are generated from [slides.Rmd](rmd_files/slides.Rmd) (this will source the contents of
content.Rmd).
* The [functions.R](rmd_files/functions.R) script is a duplicate of the one in the top directory 
and is required for knitting the various files. 
* The [render_rmakdown_files.R](rmd_files/render_rmakdown_files.R) script contains four code 
chunks. Running them will re-generate all the course material. This file should always be run before
you merge any changes to the `main` branch.

### Tips when presenting
* You can press F11 to make the presentation full screen. To go off of full screen you can press F11 again.

