---
output: rmarkdown::github_document
---

```{r setup, include=FALSE}
# This variable determines whether or not exercise solutions are included
show_solution <- FALSE
```

# Writing functions in R

This repository is for Writing Functions in R course offered by the DASD R Training Group. It was also used for the April 2019 [Coffee & Coding](https://github.com/moj-analytical-services/Coffee-and-Coding) session.

Knowing how to write your own functions is a great skill to add to your R toolbox. Writing functions can save you time, reduce the risk of errors, and make your code easier to understand. In this course we cover why, when and how to write your own functions, with plenty of examples and exercises to help you get started.

The session is intended to be accessible to anyone who has attended the [Introduction to R](https://github.com/moj-analytical-services/IntroRTraining) training session and used R a little in their work.

All the notes for the training session are available below. If you have any feedback on the content, please get in touch!

## Contents

* [Pre-material](#pre-material)
* [Learning outcomes](#learning-outcomes)
* [What is a function?](#what-is-a-function)
* [Why use functions?](#why-use-functions)
* [How to write a function](#how-to-write-a-function)
* [Examples of basic functions](#examples-of-basic-functions)
* ["Real-world" example functions](#real-world-example-functions)
* [When to write a function](#when-to-write-a-function)
* [Best practice](#best-practice)
* [How to organise your code](#how-to-organise-your-code)
* [Writing a package](#writing-a-package)
* [Further reading](#further-reading)


## Pre-material

If you want to try running the example code during or after the session, you can clone this repository.

+ If you have used GitHub and the Analytical Platform before, follow the instructions in [section 4.2.1 of the Platform User Guidance](https://moj-analytical-services.github.io/platform_user_guidance/using-github-with-the-platform.html#step-1-navigate-to-your-platform-r-studio-and-make-a-copy-of-the-github-project-in-your-r-studio).

+ If you have not used GitHub with the analytical platform before, you might need to set this up using the instructions in [section 2.3 of the Platform User Guidance](https://moj-analytical-services.github.io/platform_user_guidance/using-github-with-the-platform.html), before following the instructions in section 4.2.1 as above.

All the examples in the presentation/README are available in the R script example_code.R.

```{r source_content, child = 'content.Rmd'}
```

```{r source_content, child = 'appendix.Rmd'}
```
