---
title: Writing functions in R
output: rmarkdown::github_document
---

```{r setup, include=FALSE}
# This variable determines whether or not exercise solutions are included
show_solution <- FALSE
```

This repository is for Writing Functions in R course offered by the DASD R Training Group. It was also used for the April 2019 [Coffee & Coding](https://github.com/moj-analytical-services/Coffee-and-Coding) session.

Knowing how to write your own functions is a great skill to add to your R toolbox. Writing functions can save you time, reduce the risk of errors, and make your code easier to understand. In this course we’ll cover why, when and how to write your own functions, with plenty of examples & exercises to help you get started.

The session is intended to be accessible to anyone who is familiar with the content of the [Introduction to R](https://github.com/moj-analytical-services/IntroRTraining) training session & has some experience of using R in their work.

This training session is run in person/over Skype every few months. Alternatively, you can go through this material in your own time - all the notes are available below and you can also find the recording of a previous session [here](https://web.microsoftstream.com/channel/aa3cda5d-99d6-4e9d-ac5e-6548dd55f52a). See [Remote learning](#remote-learning) for more tips on going through this material in your own time.

If you have any feedback on the content, please get in touch!

## Contents

* [Pre-material](#pre-material)
* [Remote learning](#remote-learning)
* [Learning outcomes](#learning-outcomes)
* [What is a function?](#what-is-a-function)
* [Why use functions?](#why-use-functions)
* [How to write a function](#how-to-write-a-function)
* [Examples of basic functions](#examples-of-basic-functions)
* ["Real-world" example functions](#real-world-example-functions)
* [When to write a function](#when-to-write-a-function)
* [Best practice](#best-practice)
* [How to organise your code](#how-to-organise-your-code)
* [Writing a package](#writing-a-package)
* [Further reading](#further-reading)

## Pre-material

A few days before the session, please make sure that -

1. You have access to RStudio on the Analytical Platform
2. You have access to the [alpha-r-training s3 bucket](https://cpanel-master.services.alpha.mojanalytics.xyz/datasources/607/)
3. You have followed the steps in the [Configure Git and Github section of the Platform User Guidance](https://user-guidance.services.alpha.mojanalytics.xyz/introduction.html#configure-git-and-github) to configure Git and GitHub (this only needs doing once)
4. You have cloned this repository (instructions are in the Analytical Platform User Guidance [here](https://user-guidance.services.alpha.mojanalytics.xyz/github.html#creating-your-project-repo-on-github))

If you have any problems with the above please get in touch with the course organisers or ask for help on either the #analytical_platform or #intro_r channel on [ASD slack](https://asdslack.slack.com).

All the examples in the presentation/README are available in the R script example_code.R.

## Remote learning

Here are a few suggestions if you are going through this material in your own time:

* Both the read me and slides contain the same content so you can use whichever works best for you - the only difference is the exercise solutions are only in the slides
*  To open the slides, first clone the repo (see [pre-material](#pre-material)) and then open the file "slides.html" from RStudio by clicking "View in web browser"
* You can find a recording of a previous training session on the R training Microsoft Stream channel [here](https://web.microsoftstream.com/channel/aa3cda5d-99d6-4e9d-ac5e-6548dd55f52a)
* If you need any help, the best place to ask is on either the [#intro_r](https://app.slack.com/client/T1PU1AP6D/CGKSJV9HN) or [#r](https://app.slack.com/client/T1PU1AP6D/C1PUCG719) slack channels on ASD slack.

If you have any feedback on the material and/or your experience of working through it in your own time, please get in touch with a member of the R training group or leave a review on [Airtable](https://airtable.com/shr9u2OJB2pW8Y0Af).


```{r source_content, child = 'content.Rmd'}
```

```{r source_content, child = 'appendix.Rmd'}
```
