---
title: Writing functions in R
output: rmarkdown::github_document
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = "../",
      output_file = file.path("../README.md")  
    )
  })
---

```{r setup, include=FALSE}
# These variables determine whether or not exercise solutions are included
show_solution <- FALSE # This determines if the solutions are displayed in the readme
purl_solutions <- FALSE # This variable relates to code blocks that are exercise solutions
purl_example_code <- TRUE # This variable relates to code blocks that aren't exercise solutions
knitr::opts_chunk$set(
  comment = "#>",
  fig.path = "../README_files/"
)
```

This repository is for the Writing Functions in R course offered by the Data and Analysis R Training Group. 

Knowing how to write your own functions is a great skill to add to your R toolbox. Writing functions can save you time, reduce the risk of errors, and make your code easier to understand. In this course we’ll cover why, when and how to write your own functions, with plenty of examples & exercises to help you get started.

The session is intended to be accessible to anyone who is familiar with the content of the [Introduction to R](https://github.com/moj-analytical-services/IntroRTraining) training session & has some experience of using R in their work.

This training session is periodically run in person/over Teams. Alternatively, you can go through this material in your own time - all the notes are available below and you can also find the recording of a previous session [here](https://justiceuk.sharepoint.com/:f:/s/RTrainingGroup/Ev1kwXxGaVNNpLwFg4drakUBNUpZz72DRdcD8PipkU-u9A?e=SQzeJL). See [Remote learning](#remote-learning) for more tips on going through this material in your own time. If you work through the material by yourself please leave feedback about the material [here](https://airtable.com/shr9u2OJB2pW8Y0Af).

If you have any feedback on the content, please get in touch!

## Contents

* [Set up instructions](#set-up-instructions)
* [Remote learning](#remote-learning)
* [Learning outcomes](#learning-outcomes)
* [What is a function?](#what-is-a-function)
* [Why use functions?](#why-use-functions)
* [How to write a function](#how-to-write-a-function)
* [Examples of function-writing patterns](#examples-of-function-writing-patterns)
* ["Real-world" example functions](#real-world-example-functions)
* [When to write a function](#when-to-write-a-function)
* [Best practice](#best-practice)
* [How to organise your code](#how-to-organise-your-code)
* [Writing a package](#writing-a-package)
* [Further reading](#further-reading)
* [Appendix](#appendix)
* [Information for presenters](#information-for-presenters)

### Set up instructions

A few days before the session, please:

* Ensure you have requested access to the **alpha-r-training** s3 bucket via the [intro_r slack 
channel on the ASD workspace](https://moj.enterprise.slack.com/archives/CGKSJV9HN)
* [Ensure you have access to the Analytical Platform](https://user-guidance.analytical-platform.service.justice.gov.uk/get-started.html#quickstart-guide).
* [Ensure you have linked RStudio and GitHub with an SSH key](https://user-guidance.analytical-platform.service.justice.gov.uk/github/set-up-github.html).

Before the session, please:

* [Clone](https://user-guidance.analytical-platform.service.justice.gov.uk/github/rstudio-git.html#work-with-git-in-rstudio) this repo as an RStudio project.
* Run `install.packages("renv")`
* Run `renv::install()` to install the packages declared in the `DESCRIPTION` file.

## Remote learning

This training session is run in person/over Teams every few months. Alternatively, you can go 
through this material in your own time. Recordings of these sessions can be viewed via links 
provided in the [Analytical Platform and related tools training section on R training](https://moj-analytical-services.github.io/ap-tools-training/ITG.html#r-training). If you 
work through the material by yourself please leave feedback about the material 
[here](https://airtable.com/shr9u2OJB2pW8Y0Af).  

**Please contact [Aidan Mews](aidan.mews@justice.gov.uk) if you have any questions.**

Here are a few suggestions if you are going through this material in your own time:

* Both the README and slides contain the same content so you can use whichever works best for you - 
the only difference is the exercise solutions are only in the slides
*  To open the slides, first clone the repo (see [set up instructions](#set-up-instructions)) and then open the 
file "slides.html" from RStudio by clicking "View in web browser"

```{r source_content, child = 'content.Rmd'}
```

```{r source_content, child = 'appendix.Rmd'}
```

## Information for presenters

### Where stuff is for on the day
* The training slides are in [slides.html](slides.html) (you will need to download these and open them in a browser).
* The course material is also duplicated in the README with some additional examples in the appendix.
* The code from the presentation is available in [example_code.R](example_code.R).
* The answers to the exercises are available in [solutions.R](solutions.R).
* The [functions.R](functions.R) script is used in one example to demonstrate how to organize code.

### How to update the course contents
The README, slides, solutions, and example_code are all generated from files in the "rmd_files" folder.

* The bulk of the content is in [rmd_files/content.Rmd](rmd_files/content.Rmd).
* The appendix (used in the README) is in [rmd_files/appendix.Rmd](rmd_files/appendix.Rmd).
* The README is generated from [rmd_files/README.Rmd](rmd_files/README.Rmd) (this will source the contents of
content.Rmd and appendix.Rmd).
* The slides are generated from [rmd_files/slides.Rmd](rmd_files/slides.Rmd) (this will source the contents of
content.Rmd). You may need to disable the large file pre-commit hook for this (see below). 
* The [rmd_files/functions.R](rmd_files/functions.R) script is a duplicate of the one in the top directory 
and is required for knitting the various files. 
* The [rmd_files/render_rmakdown_files.R](rmd_files/render_rmakdown_files.R) script contains four code 
chunks. Running them will re-generate all the course material. This file should always be run before
you merge any changes to the `main` branch.

* Disabling the large file pre-commit hook: 
    * Ensure hidden files are visible (in the `Files` tab in the bottom right Rstudio pane, select 
    the `More` cog wheel and check that `Show Hidden Files` is ticked).
    * Navigate to `.git/hooks/pre-commit`
    * Increase the file size (edit the line `FILE_SIZE_LIMIT=5  # In MB` and save)

### Tips when presenting
* You can press F11 to make the presentation full screen. To go off of full screen you can press F11 again.

