---
title: Writing functions in R
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
---

```{r setup, include=FALSE}
# These variables determine whether or not exercise solutions are included
show_solution <- TRUE # This determines if the solutions are displayed in the slides
purl_solutions <- FALSE # This variable relates to code blocks that are exercise solutions
purl_example_code <- TRUE # This variable relates to code blocks that aren't exercise solutions
```

## Writing functions in R: setup

Before we start, make sure you've:

1. Got access to the `alpha-r-training` bucket
2. Cloned the `writing_functions_in_r` repo
3. Pulled any changes (the blue arrow on the Git tab in the top right panel)
4. Got `example_code.R` open
5. Opened the `writing_functions_in_r` repo on GitHub, so you can refer to the read me

If you need any help, just ask! For anyone joining remotely, we will be monitoring Teams throughout the session.

Note we will be recording this session for upload later to the R training Microsoft Stream channel.

## Agenda

1. Set up and introductions - 10 mins
2. Examples of basic functions (including exercises) - 1 hour 10 mins
3. Break - 10 mins
4. 'Real-world' example functions (including exercises) - 1 hour 15 mins
5. Hints and tips - 15 mins


```{r source_content, child = 'content.Rmd'}
```

# Any questions?