---
title: Writing functions in R
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
---

```{r setup, include=FALSE}
# These variables determine whether or not exercise solutions are included
show_solution <- TRUE # This determines if the solutions are displayed in the slides
purl_solutions <- FALSE # This variable relates to code blocks that are exercise solutions
purl_example_code <- TRUE # This variable relates to code blocks that aren't exercise solutions
```

## Writing functions in R: setup

Before we start, make sure you've:

1. Got access to the `alpha-r-training` bucket
2. Cloned the `writing_functions_in_r` repo
3. Pulled any changes (the blue arrow on the Git tab in the top right panel)
4. Got `example_code.R` open
5. Opened the `writing_functions_in_r` repo on GitHub, so you can refer to the read me

If you need any help, just ask! We will be monitoring the Teams chat throughout the session.

## Agenda

Item                       | Time
---------------------------| -------------
Set up and introduction    | 10:00 - 10:10
Examples of basic functions (including exercises) | 10:10 - 11:20
Break | 11:20 - 11:30
'Real-world' example functions (including exercises) | 11:30 - 12:45
Hints and tips | 12:45 - 13:00

```{r source_content, child = 'content.Rmd'}
```

# Any questions?