---
title: Writing functions in R
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = "../",
      output_file = file.path("../slides.html")  
    )
  })
---

```{r setup, include=FALSE}
# These variables determine whether or not exercise solutions are included
show_solution <- TRUE # This determines if the solutions are displayed in the slides
purl_solutions <- FALSE # This variable relates to code blocks that are exercise solutions
purl_example_code <- TRUE # This variable relates to code blocks that aren't exercise solutions
```

## Writing functions in R: setup

Before we start, make sure you've:

1. Got access to the `alpha-r-training` bucket
2. Cloned the [writing_functions_in_r](https://github.com/moj-analytical-services/writing_functions_in_r) repo into your RStudio
3. Pulled any changes (the blue arrow on the Git tab in the top right panel)
4. Got `example_code.R` open in RStudio
5. Run the command `renv::restore()` in the RStudio console to make sure you have the required packages installed
6. Opened the [writing_functions_in_r](https://github.com/moj-analytical-services/writing_functions_in_r) repo on GitHub, so you can refer to the README

If you need any help, just ask! We will be monitoring the Teams chat throughout the session.

## Agenda

Item                                                 | Time
---------------------------------------------------- | -------------
Set up and introduction                              | 10:00 - 10:10
Examples of basic functions (including exercises)    | 10:10 - 11:30
Break                                                | 11:30 - 11:40
'Real-world' example functions (including exercises) | 11:40 - 12:50
Hints and tips                                       | 12:50 - 13:00

```{r source_content, child = 'content.Rmd'}
```

# Any questions?
